---
output:
  pdf_document:
    latex_engine: xelatex
header-includes:
  # Packages
  - '\usepackage{fancyhdr}'
  - '\usepackage{float}'
  - '\usepackage{etoolbox}'
  - '\usepackage{framed}'
  - '\usepackage[most]{tcolorbox}'
  # Page style
  - '\pagestyle{fancy}'
  - '\fancyhf{}'
  - '\fancyhead[R]{ID: 114078510}'
  - '\fancyfoot[R]{\footnotesize\thepage}'
  - '\renewcommand{\headrulewidth}{0pt}'
  - '\setlength{\leftmargini}{1.5em}'
  - '\setlength{\leftmarginii}{1.5em}'
  # Section and spacing
  - '\makeatletter'
  - '\renewcommand{\section}{\@startsection{section}{1}{\z@}{0.1\baselineskip}{0.1\baselineskip}{\normalfont\Large\bfseries}}'
  - '\preto{\@verbatim}{\topsep=0pt \partopsep=0pt \parskip=0pt \parsep=0pt}'
  - '\AtBeginEnvironment{itemize}{\setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}\setlength{\topsep}{0pt}}'
  - '\makeatother'
  # Custom boxes/environments
  - '\newtcolorbox{remarkbox}{colback=yellow!6,colframe=orange!55!yellow,arc=1.5mm,boxrule=0.4pt,left=4pt,right=4pt,top=2pt,bottom=1pt}'
  - '\setlength{\OuterFrameSep}{1.5pt}'
  - '\makeatletter'
  - '\@ifundefined{Shaded}{\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}\vspace{-0.2\baselineskip}}}{\renewenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}\vspace{-0.2\baselineskip}}}'
  - '\makeatother'
---

# CSDS 2026 -- HW 3
\begin{remarkbox}
\footnotesize\textit{\textbf{Remark:} Using AI to refine plots and RMarkdown formatting, and to improve the grammar of written answers.}
\end{remarkbox}

\medskip

### Initialization
```{r setup}
set.seed(20260311)
library(ggplot2)

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.pos = "H",
  fig.align = "center",
  fig.show = "hold",
  dev.args = list(pointsize = 9)
)
```

\begingroup\color{gray!45}\noindent\rule{\linewidth}{0.3pt}\par\endgroup

## Question 1
Let’s reexamine how to *standardize* data: subtract the mean of a vector from all its values, and divide this difference by the standard deviation to get a vector of standardized values.

\par

### (a) Create a normal distribution (mean=940, sd=190) and standardize it to `rnorm_std`

```{r q1a}
rnorm_data <- rnorm(n = 1000, mean = 940, sd = 190)
rnorm_std <- scale(rnorm_data)
```

i. **What should we expect the mean and standard deviation of `rnorm_std` to be, and why?**
    \par
    We expect the mean of `rnorm_std` to be 0 and the standard deviation to be 1, because standardization centers the data at 0 and scales it to unit variance. 
		\par
		And it did happen as expected:
    ```{r q1a_1}
    mean_std <- mean(rnorm_std)
    sd_std <- sd(rnorm_std)
    cat("Mean:", mean_std, "\nStandard Deviation:", sd_std)
    ```

\par

ii. **What should the distribution (shape) of `rnorm_std` look like, and why?**
    \par
    The distribution of `rnorm_std` should look like a standard normal distribution (bell-shaped curve) because standardization does not change the shape of the distribution, it only shifts and scales it.
    ```{r q1a_2, out.width='75%'}
    # --- predefine format for the plot ---
    theme_set(theme_minimal())

    base_theme <- theme(
      plot.title = element_text(hjust = 0.5),
      legend.position = c(0.99, 0.99),
      legend.justification = c(1, 1),
      legend.title = element_blank()
    )

		plot_density <- function(
			x,
			title,
			x_label,
			fill_color = "#cee9f2",
			line_color = "#3e5ed2",
			show_hist = TRUE
		) {
			p <- ggplot(data.frame(value = x), aes(x = .data[["value"]]))

			if (show_hist) {
				p <- p + geom_histogram(
					aes(y = after_stat(density)), fill = fill_color, color = "white"
				)
			}

			p + geom_density(color = line_color, linewidth = 1) +
			labs(title = title, x = x_label, y = "Density") + base_theme
		}
    # -------------------------------------

    plot_density(
      x = rnorm_std,
      title = "Histogram of rnorm_std",
      x_label = "Standardized Values"
    )
    ```

\par

iii. **What do we generally call distributions that are normal and standardized?**
    \par
		\smallskip
    We generally call them "standard normal distribution" or "Z-distribution", which has a mean of 0 and a standard deviation of 1.

\medskip

### (b) Create a standardized version of `minday` discussed in question 3 (let’s call it `minday_std`)

```{r q1b}
# minday from question 3
bookings <- read.table("first_bookings_datetime_sample.txt", header = TRUE)
hours <- as.POSIXlt(bookings$datetime, format = "%m/%d/%Y %H:%M")$hour
mins <- as.POSIXlt(bookings$datetime, format = "%m/%d/%Y %H:%M")$min
minday <- hours * 60 + mins

# create minday_std
minday_std <- scale(minday)
```

i. **What should we expect the mean and standard deviation of `minday_std` to be, and why?**
    \par
    We should expect the mean of `minday_std` to be 0 and the standard deviation to be 1, because standardization centers the data at 0 and scales it to unit variance, same as with `rnorm_std`.
		\par
		And it did happen as expected:
    ```{r q1b_1}
    mean_minday_std <- mean(minday_std)
    sd_minday_std <- sd(minday_std)
    cat("Mean:", mean_minday_std, "\nStandard Deviation:", sd_minday_std)
    ```

\medskip

ii. **What should the distribution of `minday_std` look like compared to `minday`, and why?**
    \par
    The distribution of `minday_std` should have the same shape as `minday`, including skewness, peaks, and other characteristics, but shifted to have a mean of 0 and scaled to have a standard deviation of 1.

    ```{r q1b_2_side_by_side, fig.show='hold', fig.width=5, fig.height=4, out.width='49%', fig.align='center'}
    plot_density(
      x = minday, title = "Histogram of minday",
      x_label = "Values", fill_color = "#f2d9ce", line_color = "#d23e3e"
    )

    plot_density(
      x = minday_std, title = "Histogram of minday_std",
      x_label = "Standardized Values"
    )
    ```


## Question 2
Install the “compstatslib” package from Github (see week 2 class handout) and run the `plot_sample_ci()` function that simulates samples drawn randomly from a population.

\par

### (a) Simulate 100 samples (each of size 100), from a normally distributed population of 10,000

```{r q2a, out.width='70%'}
library(compstatslib)

plot_sample_ci(num_samples = 100, sample_size = 100, pop_size = 10000,
               distr_func = rnorm, mean = 20, sd = 3)
title(main = "(a) Normally Distributed, n = 100")
```

\par

i. **How many samples do we expect to NOT include the population mean in its 95% CI?**
    \par
    We expect about `100 * 0.05 = 5` samples to not include the population mean in their 95% confidence intervals.

\smallskip

ii. **How many samples do we expect to NOT include the population mean in their 99% CI?**
    \par
    We expect about `100 * 0.01 = 1` sample to not include the population mean in their 99% confidence intervals.

\medskip

### (b) Rerun the previous simulation with the same number of samples, but larger sample size (`sample_size`=300):

```{r q2b, out.width='70%'}
plot_sample_ci(num_samples = 100, sample_size = 300, pop_size = 10000,
               distr_func = rnorm, mean = 20, sd = 3)
title(main = "(b) Normally Distributed, n = 300")
```

i. **Now that the size of each sample has increased, do we expect their 95% and 99% CI to become wider or narrower than before?**
    \par
    The confidence interval is given by Eq.\ \eqref{eq:ci-mean}.

    \begin{equation}
    \label{eq:ci-mean}
    \bar{x} \pm z \frac{\sigma}{\sqrt{n}}
    \end{equation}

    As the sample size \(n\) increases, the standard error \( \frac{\sigma}{\sqrt{n}} \) becomes smaller. So both the 95% and 99% confidence intervals become **narrower** than before.

\smallskip

ii. **This time, how many samples (out of the 100) would we expect to NOT include the population mean in its 95% CI?**
    \par
    We still expect about `100 * 0.05 = 5` samples to not include the population mean in their 95% confidence intervals, because the confidence level (95%) has not changed.

\medskip

### (c) If we ran the above two examples (a and b) using a uniformly distributed population, how do you *expect* your answers to (a) and (b) to change, and *why*?
```{r q2c, fig.show='hold', out.width='49%', fig.align='center'}
# (a)
plot_sample_ci(num_samples = 100, sample_size = 100, pop_size = 10000,
               distr_func = runif, min = 0, max = 40)
title(main = "(a) Uniformly Distributed, n = 100")

# (b)
plot_sample_ci(num_samples = 100, sample_size = 300, pop_size = 10000,
               distr_func = runif, min = 0, max = 40)
title(main = "(b) Uniformly Distributed, n = 300")
```
Compared with the case of *normally distributed* population, the expected miss counts are still about the same, with around 5 out of 100 for 95% CI and 1 out of 100 for 99% CI.
Although `rnorm` generates bell-shaped data and `runif` generates flat, equally likely data, CI coverage is determined by the confidence level, so changing the population shape does not change the target coverage.

\medskip

## Question 3
The company EZTABLE has an online restaurant reservation platform that is accessible by mobile and web. Imagine that EZTABLE would like to start a promotion for new members to make their bookings earlier in the day.
\par
\begingroup\color{gray} A *sample* data about their \underline{new members}: the date and time for which they make their first ever booking using EZTABLE.\par\endgroup

<!-- ```{r q3a, out.width='70%'}
plot(density(minday), main = "Minute (of the day) of first ever booking",
     col = "#3e5ed2", lwd = 2)
``` -->

\medskip

### (a) What is the *average* booking time for new members making their first restaurant booking?
\vspace{-0.4\baselineskip}
\begingroup\color{gray}Note: use `minday`, which is the absolute minute of the day from 0-1440.\par\endgroup

i. **\underline{Use traditional statistical methods} to estimate the population mean of minday, its *standard error*, and the *95% confidence interval* (CI) of the sampling means**
    ```{r q3a_1}
    # We've loaded the data in Q1(b), so we can directly use `minday` here.
    minday_mean <- mean(minday)
    minday_se <- sd(minday) / sqrt(length(minday))
    minday_ci_95 <- minday_mean + c(-1.96, 1.96) * minday_se
    cat("Mean:", minday_mean, "\nStandard Error:", minday_se,
        "\n95% CI:", round(minday_ci_95[1], 2), " ~ ", round(minday_ci_95[2], 2))
    ```

\par

ii. **Bootstrap to produce 2000 new samples from the original sample**
	```{r q3a_2}
	resamples <- replicate(2000, sample(minday, length(minday), replace = TRUE))
	cat("Resamples Dimension:", dim(resamples))
	```

\par

iii. **Visualize the means of the 2000 minday bootstrapped samples**
		```{r q3a_3_1}
		resampled_means <- colMeans(resamples)
		cat("Mean of Bootstrap Sample Means:", mean(resampled_means))
		```

		\par
		We can view the means of the 2000 bootstrapped samples in 2 ways:
		\par

		(1) Showing all 2000 bootstrapped sample means as vertical lines, highlighting their average, and comparing it with the original sample mean on the `minday` density plot.
		```{r q3a_3, out.width='80%', fig.width=7, fig.height=4}
		# ----- predefine function for the plot -----
		plot_sampling_center_lines <- function(
			x,
			centers,
			title,
			center_label,
			x_label = "Value",
			line_color = "#e0582a"
		) {
			plot_density(x = x, title = title, x_label = x_label) +
				geom_vline(
					aes(xintercept = mean(centers), color = center_label),
					linewidth = 1.3, linetype = "dashed",
					key_glyph = "path"
				) +
				geom_vline(
					xintercept = centers, color = line_color,
					linewidth = 0.3, alpha = 0.25
				) +
				scale_color_manual(values = setNames(line_color, center_label)) +
				base_theme
		}
		# -------------------------------------

		plot_sampling_center_lines(
			x = minday,
			centers = resampled_means,
			center_label = "Bootstrap Mean",
			title = "Minute (of the day) of first ever booking",
			x_label = "Minute"
		)
		```

		(2) Simply observe the distribution of the bootstrapped means.
		```{r q3a_3_3, out.width='80%', fig.width=7, fig.height=4}
		plot_density(
			x = resampled_means,
			title = "Distribution of Bootstrap Means",
			x_label = "Bootstrap Mean",
			show_hist = FALSE,
			line_color = "#e0582a"
		)
		```

\par

iv. **Estimate the *95% CI* of the bootstrapped means \underline{using the quantile function}**
    ```{r q3a_4}
    quantile(resampled_means, probs = c(0.025, 0.975))
    ```

\medskip

### (b) By what time of day, have half the new members of the day already arrived at their restaurant?

i. **Estimate the *median* of `minday`**
    ```{r q3b_1}
    minday_median <- median(minday)
    cat("Median:", minday_median)
    ```

\par

ii. **Visualize the *medians* of the 2000 bootstrapped samples**
	```{r q3b_2_1}
	resampled_medians <- apply(resamples, 2, median)
	cat("Mean of Bootstrap Sample Medians:", mean(resampled_medians))
	```

	\smallskip

	\par
	We can view the medians of the 2000 bootstrapped samples in 2 ways:
	\par

	(1) Showing all 2000 bootstrapped sample medians as vertical lines, highlighting their average, and comparing it with the original sample median on the `minday` density plot.
	```{r q3b_2, out.width='80%', fig.width=7, fig.height=4}
	plot_sampling_center_lines(
		x = minday,
		centers = resampled_medians,
		center_label = "Bootstrap Median",
		title = "Minute (of the day) of first ever booking",
		x_label = "Minute",
		line_color = "#627b08"
	)
	```

	(2) Simply observe the distribution of the bootstrapped medians.
	```{r q3b_2_3, out.width='80%', fig.width=7, fig.height=4}
	plot_density(
		x = resampled_medians,
		title = "Distribution of Bootstrap Medians",
		x_label = "Bootstrap Median",
		show_hist = FALSE,
		line_color = "#627b08"
	)
	```

\par

iii. **Estimate the *95% CI* of the bootstrapped medians \underline{using the quantile function}**
      ```{r q3b_3}
      quantile(resampled_medians, probs = c(0.025, 0.975))
      ```
