---
output:
  pdf_document:
    latex_engine: xelatex
header-includes:
  # Packages
  - '\usepackage{fancyhdr}'
  - '\usepackage{float}'
  - '\usepackage{etoolbox}'
  - '\usepackage{framed}'
  - '\usepackage[most]{tcolorbox}'
  # Page style
  - '\pagestyle{fancy}'
  - '\fancyhf{}'
  - '\fancyhead[R]{ID: 114078510}'
  - '\fancyfoot[R]{\footnotesize\thepage}'
  - '\renewcommand{\headrulewidth}{0pt}'
  - '\setlength{\leftmargini}{1.5em}'
  - '\setlength{\leftmarginii}{1.5em}'
  # Section and spacing
  - '\makeatletter'
  - '\renewcommand{\section}{\@startsection{section}{1}{\z@}{0.1\baselineskip}{0.1\baselineskip}{\normalfont\Large\bfseries}}'
  - '\preto{\@verbatim}{\topsep=0pt \partopsep=0pt \parskip=0pt \parsep=0pt}'
  - '\AtBeginEnvironment{itemize}{\setlength{\itemsep}{1.2em}\setlength{\parskip}{0pt}\setlength{\topsep}{0.2em}}'
  - '\makeatother'
  # Custom boxes/environments
  - '\newtcolorbox{remarkbox}{colback=yellow!6,colframe=orange!55!yellow,arc=1.5mm,boxrule=0.4pt,left=4pt,right=4pt,top=2pt,bottom=1pt}'
  - '\newtcolorbox{recallbox}{colback=gray!6,colframe=black!60,arc=1.5mm,boxrule=0.4pt,left=4pt,right=4pt,top=2pt,bottom=1pt}'
  - '\setlength{\OuterFrameSep}{1.5pt}'
  - '\makeatletter'
  - '\@ifundefined{Shaded}{\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}\vspace{-0.2\baselineskip}}}{\renewenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}\vspace{-0.2\baselineskip}}}'
  - '\makeatother'
---

# CSDS 2026 -- Week 9
\begin{remarkbox}
\footnotesize\textit{\textbf{Remark:} Using AI to improve the table format, the grammar of answers and multicollinearity explanation for Q3(c).}
\end{remarkbox}

\medskip

### Initialization
```{r setup}
set.seed(20260424)
library(compstatslib)
```

\begingroup\color{gray!45}\noindent\rule{\linewidth}{0.3pt}\par\endgroup

## Question 1
We will use the `interactive_regression()` function from CompStatsLib again

- Scenario 1: Consider a very narrowly dispersed set of points that have a negative or positive steep slope 
- Scenario 2: Consider a widely dispersed set of points that have a negative or positive steep slope 
- Scenario 3: Consider a very narrowly dispersed set of points that have a negative or positive shallow slope 
- Scenario 4: Consider a widely dispersed set of points that have a negative or positive shallow slope

```{r question1, fig.width=10, fig.height=3}
# pts1 <- interactive_regression()
# pts2 <- interactive_regression()
# pts3 <- interactive_regression()
# pts4 <- interactive_regression()

# all_pts <- list(
#   scenario1 = pts1,
#   scenario2 = pts2,
#   scenario3 = pts3,
#   scenario4 = pts4
# )

# saveRDS(all_pts, "all_pts.rds")

all_pts <- readRDS("all_pts.rds")
scenario1 <- all_pts$scenario1
scenario2 <- all_pts$scenario2
scenario3 <- all_pts$scenario3
scenario4 <- all_pts$scenario4

par(mfrow = c(1, 4), mar = c(3, 3, 2, 1))
plot_regr(scenario1, stats = FALSE)
plot_regr(scenario2, stats = FALSE)
plot_regr(scenario3, stats = FALSE)
plot_regr(scenario4, stats = FALSE)
```

### a. Comparing scenarios 1 and 2, which do we expect to have a stronger R2 ?
I expect `Secnario 1` to have a stronger R2 than `Scenario 2` because the points are more closely clustered around the regression line, indicating a stronger linear relationship between x and y.
\par
The results also confirm this expectation:
```{r question1a}
s1_r2 <- summary(lm(y ~ x, data = scenario1))$r.squared
s2_r2 <- summary(lm(y ~ x, data = scenario2))$r.squared
cat("Scenario 1 R2:", s1_r2, "\n")
cat("Scenario 2 R2:", s2_r2)
```

### b. Comparing scenarios 3 and 4, which do we expect to have a stronger R2 ?
I expect `Secnario 3` to have a stronger R2 than `Scenario 4` because the points are more closely clustered around the regression line, indicating a stronger linear relationship between x and y.
\par
The results also confirm this expectation:
```{r question1b}
s3_r2 <- summary(lm(y ~ x, data = scenario3))$r.squared
s4_r2 <- summary(lm(y ~ x, data = scenario4))$r.squared
cat("Scenario 3 R2:", s3_r2, "\n")
cat("Scenario 4 R2:", s4_r2)
```

### c. Comparing scenarios 1 and 2, which do we expect has bigger/smaller SSE, SSR, and SST? (intuitively)
I expect:

- SSE: ***`Scenario 1` will have a smaller SSE*** than `Scenario 2` because the points are more closely clustered around the regression line, resulting in smaller residuals.
- SSR: In this case, we expect that ***`Scenario 1` has a slightly larger SSR*** than `Scenario 2` because its fitted line explains somewhat more of the variation in y. However, since both scenarios show a steep positive linear trend, their SSR values should not differ by very much. The fitted lines pass through roughly the similar lower-left and upper-right regions, so the amount of variation explained by the regression line is fairly similar in both cases.
- SST: I expect SST would likely be ***slightly larger in `Scenario 2`*** because SST measures the total variation in the observed y values. Although both scenarios have similar fitted trends, `Scenario 2` shows much greater scatter around the regression line, which increases the overall variability of the data.

### d. Comparing scenarios 3 and 4, which do we expect has bigger/smaller SSE, SSR, and SST? (intuitively)
I expect:

- SSE: ***`Scenario 3` will have a smaller SSE*** than `Scenario 4` because the points are more closely clustered around the regression line, resulting in smaller residuals.
- SSR: `Scenarios 3` and `4` would likely have ***similar SSR*** values because both show a shallow linear trend
- SST: I expect SST would be ***larger in `Scenario 4`*** because SST measures the total variation in the observed y values. Although both scenarios have similar fitted trends, `Scenario 4` shows much greater scatter around the regression line, which increases the overall variability of the data.


## Question 2
Let’s analyze the `programmer_salaries.txt` dataset we saw in class.  
\par
Read the file using `read.csv("programmer_salaries.txt", sep="\t")` because the columns are separated by tabs `(\t)`.
\smallskip
```{r question2}
library(data.table)
salaries <- fread("programmer_salaries.txt", sep = "\t")
head(salaries)
```

### a. Use the `lm()` function to estimate the regression model `Salary ~ Experience + Score + Degree`. Show the `beta coefficients`, `R2`, and the first 5 values of `fitted.values` and `residuals`
```{r question2a_1}
model <- lm(Salary ~ Experience + Score + Degree, data = salaries)

# Beta Coefficients
model$coefficients
```
```{r question2a_2}
# R Squared
summary(model)$r.squared
```

```{r question2a_3}
# First 5 Fitted Values
fitted(model)[1:5]
```

```{r question2a_4}
# First 5 Residuals
resid(model)[1:5]
```

### b. Use only linear algebra and the geometric view of regression to estimate the regression yourself: 

- i. **Create an `X` matrix that has a first column of 1s followed by columns of the independent variables *(only show the code)***
  ```{r question2b_1}
  X <- as.matrix(cbind(
    Intercept = 1,
    Experience = salaries$Experience,
    Score = salaries$Score,
    Degree = salaries$Degree
  ))
  ```

- ii. **Create a `y` vector with the Salary values *(only show the code)***
  ```{r question2b_2}
  y <- as.matrix(salaries$Salary)
  ```

- iii. **Compute the `beta_hat` vector of estimated regression coefficients *(show the code and values)***
  \par
  \begin{equation}
  \hat{\beta} = (X^T X)^{-1} X^T y
  \label{eq:beta_hat}
  \end{equation}
  ```{r question2b_3}
  beta_hat <- solve(t(X) %*% X) %*% t(X) %*% y
  beta_hat
  ```

- iv. **Using the above, compute a `y_hat` vector of estimated values, and a `res` vector of residuals *(show the code and the first 5 values of `y_hat` and `res`)***
  \par
  \begin{equation}
  \hat{y} = X \hat{\beta}
  \label{eq:y_hat}
  \end{equation}
  \begin{equation}  \varepsilon = y - \hat{y}
  \label{eq:residuals}
  \end{equation}
  ```{r question2b_4}
  y_hat <- X %*% beta_hat
  res <- y - y_hat
  
  # First 5 Estimated Values
  y_hat[1:5]
  ```
  ```{r question2b_4_2}
  # First 5 Residuals
  res[1:5]
  ```

- v. **Using only the results from (i) – (iv), compute SSR, SSE and SST *(show the code and values)***
  \par
  \begin{equation}
  SSR = \sum_{i=1}^{n} (\hat{y}_i - \bar{y})^2
  \label{eq:ssr}
  \end{equation}
  \begin{equation}
  SSE = \sum_{i=1}^{n} (y_i - \hat{y}_i)^2
  \label{eq:sse}
  \end{equation}
  \begin{equation}
  SST = \sum_{i=1}^{n} (y_i - \bar{y})^2
  \label{eq:sst}
  \end{equation}
  ```{r question2b_5}
  y_bar <- mean(y)
  
  SSR <- sum((y_hat - y_bar)^2)
  SSE <- sum((y - y_hat)^2)
  SST <- sum((y - y_bar)^2)
  
  result_table <- data.frame(
    Measure = c("SSR", "SSE", "SST"),
    Value = c(SSR, SSE, SST)
  )

  knitr::kable(result_table)
  ```


### c. Compute R2 for in two ways, and confirm you get the same results *(show code and values)*: 

- i. **Use any combination of SSR, SSE, and SST**
  \par
  \begin{equation}  R^2 = \frac{SSR}{SST} = 1 - \frac{SSE}{SST}
  \label{eq:r_squared}
  \end{equation}
  ```{r question2c_1}
  r_squared_1 <- SSR / SST
  r_squared_2 <- 1 - (SSE / SST)
  ```

- ii. **Use the squared correlation of vectors `y` and `y_hat`**
  \par
  \begin{equation}
  R^2 = \operatorname{cor}(y, \hat{y})^2
  \label{eq:r_squared}
  \end{equation}
  ```{r question2c_2}
  r_squared_3 <- cor(y, y_hat)^2
  ```

Below are the results from methods (i) and (ii), which confirm that they are identical:
```{r question2c}
r2_table <- data.frame(
  Method = c(
    "SSR / SST",
    "1 - SSE / SST",
    "cor(y, y_hat)^2"
  ),
  `R_squared` = c(r_squared_1, r_squared_2, r_squared_3)
)

knitr::kable(r2_table, digits = 6)
```


## Question 3
We’re going to look back at the early days of global car manufacturing, when American, Japanese, and European cars competed to rule the world.

| No. | Variable       | Description                                          |
|:---:|----------------|------------------------------------------------------|
| 1   | `mpg`          | miles-per-gallon (dependent variable)                |
| 2   | `cylinders`    | cylinders in engine                                  |
| 3   | `displacement` | size of engine                                       |
| 4   | `horsepower`   | power of engine                                      |
| 5   | `weight`       | weight of car                                        |
| 6   | `acceleration` | acceleration ability of car                          |
| 7   | `model_year`   | year model was released                              |
| 8   | `origin`       | place car was designed (1: USA, 2: Europe, 3: Japan) |
| 9   | `car_name`     | make and model names                                 |

```{r question3}
auto <- read.table("auto-data.txt", header=FALSE, na.strings = "?")
auto <- na.omit(auto)
names(auto) <- c("mpg", "cylinders", "displacement", "horsepower", "weight",
                 "acceleration", "model_year", "origin", "car_name")
```

### a. Let’s first try exploring this data and problem:

- i. **Visualize the data as you wish to answer the following questions (report only relevant plots)**
  ```{r question3a_1_2}
  par(mfrow = c(2, 4), cex.main = 0.9)

  vars <- c("cylinders", "displacement", "horsepower", "weight",
            "acceleration", "model_year", "origin")

  for (v in vars) {
    plot(
      auto[[v]], auto$mpg,
      xlab = v, ylab = "mpg",
      main = paste("mpg vs", v),
      pch = 19, cex = 0.5
    )
  }
  ```

- ii. **Report a correlation table of all variables, rounding to two decimal places**
  ```{r question3a_2}
  numeric_vars <- auto[, sapply(auto, is.numeric)]
  correlation_matrix <- round(cor(numeric_vars), 2)
  knitr::kable(correlation_matrix)
  ```

- iii. **From the visualizations and correlations, which variables appear to relate to `mpg`?**
  \par
  Among the numeric variables, `cylinders`, `displacement`, `horsepower`, and `weight` appear to have a negative relationship with `mpg`, while `model_year` show a positive relationship. The variable `origin` also seems to have a positive relationship with `mpg`.

- iv. **Which relationships might not be linear? *(don’t worry about linearity for rest of this HW)***
  \par
  - The relationships between `mpg` and `cylinders`, `horsepower`, `weight`, and `displacement` appear to be potentially non-linear, as the scatterplots suggest some curvature.
  - The relationship with `acceleration` appears less clearly linear because the points are more widely scattered and do not follow a strong straight-line pattern.
  - Since `origin` is a categorical variable, its relationship with `mpg` is not naturally represented by a simple linear pattern.

- v. **Are there any pairs of *independent variables* that are highly correlated *(r > 0.7)*?**
  \par
  Since the question asks for highly correlated pairs and does not restrict this to positive correlations only, I consider the absolute value of the correlation and identify pairs with \(|r| > 0.7\) (though the results are all positive).
  ```{r question3a_5}
  # Exclude the dependent variable 'mpg' from the correlation matrix
  # focus on independent variables only
  predictor_vars <- numeric_vars[, colnames(numeric_vars) != "mpg"]
  predictor_corr_matrix <- round(cor(predictor_vars), 2)

  high_corr_idx <- which(
    abs(predictor_corr_matrix) > 0.7 & upper.tri(predictor_corr_matrix),
    arr.ind = TRUE
  )

  high_corr_pairs <- data.frame(
    Var1 = rownames(predictor_corr_matrix)[high_corr_idx[, 1]],
    Var2 = colnames(predictor_corr_matrix)[high_corr_idx[, 2]],
    Correlation = predictor_corr_matrix[high_corr_idx]
  )

  high_corr_pairs <- high_corr_pairs[
    order(-abs(high_corr_pairs$Correlation)),
  ]

  knitr::kable(high_corr_pairs)
  ```

### b. Let’s create a linear regression model where mpg is dependent upon all other suitable variables
\textcolor{gray!70}{
  \textit{\textbf{Note:} `origin` is categorical with three levels, so use `factor(origin)` in `lm(...)` to split it into two dummy variables}
}

- i. **Which independent variables have a ‘significant’ relationship with mpg at 1% significance?**
  ```{r question3b_1}
  model_mpg <- lm(mpg ~ cylinders + displacement + horsepower + weight +
                        acceleration + model_year + factor(origin),
                  data = auto)
  summary(model_mpg)
  ```
  \medskip
  Based on the result, the independent variables that have a significant relationship with `mpg` at the 1% significance level are:
  - `displacement`
  - `weight`
  - `model_year`
  - `origin` (captured by the dummy variables created from `factor(origin)`; `factor(origin)2` represents Europe and `factor(origin)3` represents Japan, and both are statistically significant relative to the baseline category, USA)

- ii. **Looking at the coefficients, is it possible to determine which independent variables are the most effective at increasing mpg? If so, which ones, and if not, why not?**
\par
\textcolor{gray!70}{
  \textit{\textbf{hint:} units!}
}
  \par
  It is not appropriate to determine which variables are most effective at increasing mpg by comparing the raw coefficients alone, because the predictors are measured in different units and origin is a categorical variable. The coefficients can show the direction of the relationship, but not the relative importance of the variables.

### c. Let’s try to resolve some of the issues with our regression model above. 

- i. **Create *fully standardized* regression results: are these slopes easier to compare?**
\par
\textcolor{gray!70}{
  \textit{\textbf{Note:} consider if you should standardize `origin`}
}
  ```{r question3c_1}
  auto_std <- data.frame(scale(auto[1:7]), origin = factor(auto$origin))
  model_mpg_std <- lm(mpg ~ ., data = auto_std)
  summary(model_mpg_std)
  ```
  \medskip
  These slopes are easier to compare for the continuous predictors because they come from a standardized model. A larger absolute coefficient means a larger change in mpg in standard deviation units for a one standard deviation increase in that predictor, holding the others fixed. From this table, `weight` appears to have the strongest association in magnitude, followed by `model_year` and `displacement`.
  \par
  However, `origin2` and `origin3` are dummy variables, so their coefficients are not directly comparable to the standardized continuous predictors in quite the same way.

- ii. **Regress `mpg` over each nonsignificant independent variable, individually. Which ones become significant when we regress `mpg` over them individually?**
  ```{r question3c_2}
  nonsig_vars <- c("cylinders", "horsepower", "acceleration")

  for (var in nonsig_vars) {
    cat("\n====================\n")
    cat("Variable:", var, "\n")
    cat("====================\n")
    
    model_individual <- lm(as.formula(paste("mpg ~", var)), data = auto)
    print(summary(model_individual))
  }
  ```
  \medskip
  When regressing `mpg` over each of the nonsignificant independent variables (`cylinders`, `horsepower` and `acceleration`) individually, I found that every of them becomes significant at the 1% level. `cylinders` and `horsepower` have a stronger relationship with `mpg` with about 60% R-squared, while `acceleration` has a weaker relationship with `mpg` with about 10% R-squared.
  \par
  This suggests that the lack of significance in the full model for these variables may be due to multicollinearity, since these variables are highly correlated with other independent variables. As a result, their standard errors can be inflated, making it harder to detect significant relationships when other correlated predictors are included in the model.

- iii. **Plot the distribution of the residuals: are they normally distributed and centered around zero?**
  ```{r question3b_3, fig.width=9, fig.height=4}
  res <- resid(model_mpg)

  par(mfrow = c(1, 2))

  hist(res,
       probability = TRUE,
       main = "Residual Distribution",
       xlab = "Residuals",
       col = "#ecf5f6",
       border = "white")
  lines(density(res), col = "#4646f2", lwd = 2)
  abline(v = 0, col = "#e24545", lwd = 2, lty = 2)

  qqnorm(res,
         main = "Normal Q-Q Plot",
         pch = 19, cex = 0.6)
  qqline(res, col = "#e24545", lwd = 2)
  ```
  \medskip
  The residuals appear to be centered around zero, since the distribution is concentrated near zero and the Q-Q plot is roughly symmetric in the middle. However, they are not exactly normally distributed, since the Q-Q plot shows noticeable deviation in the upper tail, suggesting some right-skewness or heavier upper-tail behavior.